Refinements for Open Automata

### 2023-09-07 2021-Refinement/paper/refinementarticleV2.tex
zoom meeting

% OA
Open automata (that we abbreviate OA) were defined as a way to provide a semantics for open parameterised hierarchical labelled transition systems (abbreviated LTS). They were proposed as a theoretical foundation for parametrised automata used in verification tools and called \emph{pNets}.
An OA \cite{henrio:01299562} is similar to a classical automaton but with variables and holes. Variables make automata symbolic and allow them to encode infinite state systems. Holes enable the composition of automata: an automaton with a hole is an operator that takes another automaton as parameter and reacts to the actions it emits; the composed automaton is an automaton where the behaviour of one ``process parameter'' % 
 of the main automaton has been provided.
Due to their the generic nature, the notion of OA model is quite abstract but we already illustrated previously how to derive OAs for process algebra operators~\cite{henrio:01299562} or for component systems~\cite{AMHEEMA:2023,ameurboulifa:01526055}.

@@

An \emph{open transition} is a tuple \nmm{\OTg} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) the produced action, \(J' \subseteq J\) the holes involved in the transition, \(g \in \formulas\) the guard and \(\psi: V \to \terms\) the variable  assignments. Each \(\beta_j \in \actions\) is an action of the holes $j$,
To be well-formed, an open transition should use only variables of the automaton and variables appearing in the involved actions, formally: 
\begin{align*}
\fvars{g}&\subseteq \fvars{\alpha}\cup \bigcup_{j \in J'} \fvars{\beta_j} \cup V \\ \forall v\in V.\, \fvars{\psi(v)}&\subseteq \fvars{\alpha}\cup \bigcup_{j \in J'} \fvars{\beta_j} \cup V
\end{align*}
\end{definition}
A pair consisting of a state and a valuation is called a \emph{configuration} (of the automaton).
We use two operators to access pieces of information of the OA.

%\TODO{useful or not?}
\begin{definition}[Out-transition, Transition variables]
Let \(V\) be the variable names of a considered automaton, \(T\) its transitions and \(r\) one of its states.
\(\fOT[T]{r} \subset T\) are called the out-transitions of the state \(r\).
%\(\fIT[T]{r} \subset T\) are called the in-transitions of the state \(r\).
When the set \(T\) is clear from the context, it will be omitted.
The local variables of a transition \(\fvars{t}\) are all variables appearing in transition \(t\) except the variables of the automaton.
%\begin{align*}
% %&
%%	\fIT[T]{r} & = \mset{\OTg \in T}{s' = r} \\
%\end{align*}
%\vspace{-1cm}
\begin{mathpar}
%	\fguard{\OTg} \!=\! g \qquad
	\fOT[T]{r}  = \mset{\OTg \in T}{s = r}

@@

\begin{definition}[Hole-equal simulation]\label{def:HoleEqualSim}
Consider two OA with identical set of holes:
\(A_1 = \OA{S_1}{s_{01}}{V_1}{\sigma_{01}}{J}{T_1}\) and \(A_2 = \OA{S_2}{s_{02}}{V_2}{\sigma_{02}}{J}{T_2}\), the relation on configurations \(\mathcal{R}: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\symb{equal}\) simulation from $A_1$ to $A_2$ if the following conditions hold : 
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
 \bigsymb{\forall} {t_1} = \openrule
         {
           \beta_{1 j}^{j\in J'_{1}}, g_1,\psi_1}
         {s_1 \OTarrow {\alpha_1} s_1'} \in \fOT{s_1} .\,~~
\bigsymb{\bigexists} 
\mpar{t_{2x} \!=\! \openrule
         {
           \beta_{2 x j}^{j\in J'_{2x}}, g_{2 x},\psi_{2 x}}
         {s_2 \OTarrow {\alpha_{2 x}} s_{2 x}'}
}^{x \in X}\in \fOT{s_2}.\,~~\\ 
%\everymath{\displaystyle}\begin{array}{l}
V_1 \uplus V_2 \uplus  \fvars{t_1} \vdash \bigsymb{\bigexists} \mpar{\bigcup_x\fvars{t_{2x}}}. \hspace{7cm}
\\
 \mathcal{R}\mpar{s_1, s_2} \wedge g_1 \implies
 \operatorname*{\bigsymb{\bigvee}}_{x \in X}
{\everymath{\displaystyle}
\mpar{\begin{array}{l}
			\alpha_{2x} = \alpha_{1} \wedge \bigwedge_\subbox{j \in J} \beta_{2xj} = \beta_{1j} \nwedge\\[12pt]
			 g_{2x} \wedge\mathcal{R}\mpar{ s'_{1},s'_{2x}}\psubst{\psi_{2x} \uplus \psi_{1}}
		\end{array}} }
%\end{array} 
\end{multline*}
\item [(3)] Deadlock reduction:
\[
\forall \mpar{s_1, s_2} \!\in\! S_1 \!\times\! S_2, V_1 \uplus V_2  \!\vdash\! \mathcal{R}\mpar{s_1, s_2} \implies( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} ) \lor¢ \lnot( \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2}  )
\]
If there is a hole-equal simulation from $A_1$ to $A_2$, then we say that  $A_2$ simulates $A_1$; we  denote it $A_2\leq A_1$.
\end{definition}

@@

In the following definition we add a parameter $H$ which is the set of holes tracked by the simulation relation and adapt the definition by ignoring actions of the holes that are not in $H$.
\begin{definition}[Hole-tracking simulation]\label{Def:OA-Refinement}
For two OA\\ \(A_1 \defobject \OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}\) and \(A_2 \defobject \OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\), \(A_1\) is a simulation of \(A_2\) tracking holes \(H\), noted \(\wrel{A_1}{A_2}{H}\), with \(H \subseteq J_1 \cap J_2\), if there is a relation on configurations $\mathcal{R}: \mpar{S_1 \times S_2} \to \rformulas[V_1 \uplus V_2]$ such that:
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)
\begin{multline*}
	\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\[12pt]
		\quad \mpar{\forall x \in X, J'_{2x} \cap H = J'_1 \cap H} \nwedge\\[1pt]
		 V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\\hspace{5em} \mpar{\mathcal{R}\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj}   \nwedge\\
			 g_{2x} \wedge \mathcal{R}\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}} 
	\end{array} 
\end{multline*}
\item[(3)] Deadlock reduction:
\[
\forall \mpar{s_1, s_2} \!\in\! S_1 \!\times\! S_2, V_1 \uplus V_2  \vdash \mathcal{R}\mpar{s_1, s_2} \implies( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} ) \lor¢ \lnot( \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2}  )
\]
\end{definition}
Note that every action of the holes outside \(H\) is unconstrained according to the simulation relation.

>>>>

% OA
Open automata (that we abbreviate OA) were defined as a way to provide a semantics for open parameterised hierarchical labelled transition systems (abbreviated LTS). They were proposed as a theoretical foundation for parametrised automata used in verification tools and called \emph{pNets}.
An OA \cite{henrio:01299562} is similar to a classical automaton but with variables and holes. Variables make automata symbolic and allow them to encode infinite-state systems. Holes enable the composition of automata: an automaton with a hole is an operator that takes another automaton as parameter and reacts to the actions it emits; the composed automaton is an automaton where the behaviour of one ``process parameter'' % 
 of the main automaton has been provided.
Due to their the generic nature, the notion of OA model is quite abstract but we already illustrated previously how to derive OAs for process algebra operators~\cite{henrio:01299562} or for component systems~\cite{AMHEEMA:2023,ameurboulifa:01526055}.

@@

An \emph{open transition} is a structure \nmm{\OTg} made of several composing entities, equivalent to a tuple. In an open transition \(s, s' \in S\) are the source and target states, \(\alpha \in \actions\) is the resulting action that can be observed from the outside, \(J' \subseteq J\) are the holes involved in the transition, \(g \in \formulas\) is the guard that may constraint the transition, and \(\psi: V \to \terms\) are the variable  assignments that have an effect on the state of the automaton. Each \(\beta_j \in \actions\) is an action of the holes $j$,
To be well-formed, an open transition should use only variables of the automaton and variables appearing in the involved actions, formally: 
\begin{align*}
\fvars{g}&\subseteq \fvars{\alpha}\cup \bigcup_{j \in J'} \fvars{\beta_j} \cup V \\ \forall v\in V.\, \fvars{\psi(v)}&\subseteq \fvars{\alpha}\cup \bigcup_{j \in J'} \fvars{\beta_j} \cup V
\end{align*}
\end{definition}
A pair consisting of a state and a valuation is called a \emph{configuration} (of the automaton).
We use two operators to access pieces of information of the OA.

%\TODO{useful or not?}
\begin{definition}[Out-transition, Transition variables]
Let $\OAg$ be an automaton and let $r$ be a state in  $S$.
%Let \(V\) be the set of variable names of a considered automaton, \(T\) the set of transitions and \(r\) one of its states.
\(\fOT[T]{r} \subset T\) are the transition outgoing from state $r$\footnote{When the set \(T\) is clear from the context, it will be omitted and we will use $\fOT{r}$}.
%\(\fIT[T]{r} \subset T\) are called the in-transitions of the state \(r\).
The local variables of a transition \(\fvars{t}\) are all variables appearing in transition \(t\) except the variables of the automaton.
Outgoing transitions and variables are formally defined as follows.
%\begin{align*}
% %&
%%	\fIT[T]{r} & = \mset{\OTg \in T}{s' = r} \\
%\end{align*}
%\vspace{-1cm}
\begin{mathpar}
%	\fguard{\OTg} \!=\! g \qquad
	\fOT[T]{r}  = \mset{\OTg \in T}{s = r}

@@

\begin{definition}[Hole-equal simulation]\label{def:HoleEqualSim}
Consider two OA with identical set of holes:
\(A_1 = \OA{S_1}{s_{01}}{V_1}{\sigma_{01}}{J}{T_1}\) and \(A_2 = \OA{S_2}{s_{02}}{V_2}{\sigma_{02}}{J}{T_2}\), the relation on configurations \(\mathcal{R}: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\symb{equal}\) simulation from $A_1$ to $A_2$ if the following conditions hold : 
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
 \bigsymb{\forall} {t_1} = \openrule
         {
           \beta_{1 j}^{j\in J'_{1}}, g_1,\psi_1}
         {s_1 \OTarrow {\alpha_1} s_1'} \in \fOT{s_1} .\,~~
\bigsymb{\bigexists} 
\mpar{t_{2x} \!=\! \openrule
         {
           \beta_{2 x j}^{j\in J'_{2x}}, g_{2 x},\psi_{2 x}}
         {s_2 \OTarrow {\alpha_{2 x}} s_{2 x}'}
}^{x \in X}\in \fOT{s_2}.\,~~\\[2pt]
\mpar{\forall x \in X, J'_{2x} = J'_1 } \nwedge\hfill\\[1pt]
%\everymath{\displaystyle}\begin{array}{l}
V_1 \uplus V_2 \uplus  \fvars{t_1} \vdash \bigsymb{\bigexists} \mpar{\bigcup_x\fvars{t_{2x}}}. \hspace{7cm}
\\
 \mathcal{R}\mpar{s_1, s_2} \wedge g_1 \implies
 \operatorname*{\bigsymb{\bigvee}}_{x \in X}
{\everymath{\displaystyle}
\mpar{\begin{array}{l}
			\alpha_{2x} = \alpha_{1} \wedge \bigwedge_\subbox{j \in J'_1} \beta_{2xj} = \beta_{1j} \nwedge\\[12pt]
			 g_{2x} \wedge\mathcal{R}\mpar{ s'_{1},s'_{2x}}\psubst{\psi_{2x} \uplus \psi_{1}}
		\end{array}} }
%\end{array} 
\end{multline*}
\item [(3)] Deadlock reduction:
\[
\forall \mpar{s_1, s_2} \!\in\! S_1 \!\times\! S_2, V_1 \uplus V_2  \!\vdash\! \mathcal{R}\mpar{s_1, s_2} \implies( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} ) \lor¢ \lnot( \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2}  )
\]
If there is a hole-equal simulation from $A_1$ to $A_2$, then we say that  $A_2$ simulates $A_1$; we  denote it $A_2\leq A_1$.
\end{definition}

@@

In the following definition we add a parameter $H$ which is the set of holes tracked by the simulation relation and adapt the definition by ignoring actions of the holes that are not in $H$.
\begin{definition}[Hole-tracking simulation]\label{Def:OA-Refinement}
For two OA\\ \(A_1 \defobject \OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}\) and \(A_2 \defobject \OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\), \(A_1\) is a simulation of \(A_2\) tracking holes \(H\), noted \(\wrel{A_1}{A_2}{H}\), with \(H \subseteq J_1 \cap J_2\), if there is a relation on configurations $\mathcal{R}: \mpar{S_1 \times S_2} \to \rformulas[V_1 \uplus V_2]$ such that:
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)
\begin{multline*}
	\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\[12pt]
		\quad \mpar{\forall x \in X, J'_{2x} \cap H = J'_1 \cap H} \nwedge\\[1pt]
		 V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\\hspace{5em} \mpar{\mathcal{R}\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{1} \cap H} \beta_{1j} = \beta_{2xj}   \nwedge\\
			 g_{2x} \wedge \mathcal{R}\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}} 
	\end{array} 
\end{multline*}
\item[(3)] Deadlock reduction:
\[
\forall \mpar{s_1, s_2} \!\in\! S_1 \!\times\! S_2, V_1 \uplus V_2  \vdash \mathcal{R}\mpar{s_1, s_2} \implies( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} ) \lor¢ \lnot( \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2}  )
\]
\end{definition}
Note that every action of the holes outside \(H\) is unconstrained according to the simulation relation.
